Best practices for forecasting changes in biodiversity: an example using breeding birds

### 2017-09-03 manuscript/draft.Rmd
early discussion

Biodiversity forecasts are important for conservation, management, and
evaluating how well current models characterize natural systems. While the
number of forecasts for biodiversity is increasing, there is little information
available on how well these forecasts work. Most biodiversity forecasts are not
evaluated to determine how well they predict future diversity, fail to account
for uncertainty, and do not use time-series data that captures the actual
dynamics being studied. We addressed these limitations by using best practices
to explore our ability to forecast the species richness of breeding birds in
North America. We used hindcasting to evaluate six different modeling approaches
for predicting richness. Hindcasts for each method were evaluated annually for a
decade at `r numbers$N_predict_sites` sites distributed throughout the
continental United States. While each model could explain most of the variance
in richness, none of them consistently outperformed a baseline model that
predicted constant richness at each site. In particular, we found no evidence
that current methods (such as species distribution models) can successfully turn
spatial data into useful temporal predictions about biodiversity at decadale
time-scales. The best practices implemented in this study directly influence the
forecasts, the relative performance of different modeling approaches, and the
conclusions about the current state of biodiversity forecasting . To facilitate
the rapid improvement of biodiversity forecasts, we emphasize the value of
specific best practices in making forecasts and evaluating forecasting methods.

@@

Forecasting is an emerging imperative in ecology; as such, the field needs to
develop and follow best practices for conducting and evaluating ecological
forecasts [@clark2001]. We have used a number of these practices (Box 1) in a
single study that builds and evaluates forecasts of biodiversity in the form of
species richness. The results of this effort are both promising and
humbling. When comparing forecasts across sites, many different approaches to
forecasting produce reasonable forecasts (Figure 3). If a site is predicted to
have a high number of species in the future, relative to other sites, it
generally does. However, none of the methods evaluated reliably determined how
site-level richness changes over time (Figure 6), which is generally the stated
purpose of these forecasts. As a result, baseline models, which did not attempt
to anticipate changes in richness over time, generally provided the best
forecasts for future biodiversity. It should be noted that the scope of
inference for our results is limited to local scale forecasts for breeding birds
in North America. As such we cannot draw meaningful conclusions about global
scale forecasts of biodiversity loss or extinction rates, or about about
forecasting biodiversity changes in areas of the world like the tropics that are
expected to be heavily impacted by global change.

The most commonly used methods for forecasting future biodiversity, SDMs and
macroecological models, produced worse forecasts than time-series models and
simple baselines. This highlights the need for using hindcasting and time-series
data for assessing the effectiveness of forecasting methods and suggests that
predictions about future biodiversity change that have not been validated in
this way should be viewed with skepticism. Since site-level richness is
relatively stable, model evaluations that show high accuracy across spatial
gradients may not indicate that a model is able to predict changes over
time. This issue has also recently been highlighted as a potential downside of
using SDMs to predict the future locations of individual species
[@rapacciuolo2012; @oedekoven2017]. This result is particularly sobering because
SDMs are one of the main foundations for estimates of the predicted loss of
biodiversity to climate change [@thomas2004; @thuiller2011; @urban2015]. These
results also highlight the importance of comparing multiple modeling approaches
when conducting ecological forecasts, and in particular, the value of comparing
results to simple baselines to avoid over-interpreting the information present
in these forecasts [Box 1]. Disciplines that have more mature forecasting
cultures often do this by reporting "forecast skill", i.e., the improvement in
the forecast relative to a simple baseline [@jolliffe2003]. We recommend
following the example of @ye2015 and adopting this approach in future ecological
forecasting research.

When comparing different methods for forecasting our results demonstrate the
importance of considering uncertainty [Box 1; @clark2001; @dietz2016]. Previous
comparisons between stacked SDMs and macroecological models reported that the
methods yielded equivalent results for forecasting diversity [@algar2009;
@distler2015]. While our results support this equivalence for point estimates,
they also show that stacked SDMs dramatically underestimate uncertainty. For
this reason, direct modeling of richness is a better overall approach to
forecasting richness in this case. A similar result is seen when comparing joint
species distribution models (JSDMs) to stacked single species distribution
models. For point estimates, the joint distribution models are roughly
equivalent to stacked SDMs, but the JSDMs provide better estimates of
uncertainty [@harris2015; @warton2015]. In fact, JSDMs and direct richness
modeling provide some of the best estimates of uncertainty across all modeling
approaches. Failing to consider uncertainty in model comparisons can result
changes in the choice of methods and can lead to over confident models, which
could result in predictions that biodiversity would be restricted to a narrower
range than would actually occur.

>>>>

Biodiversity forecasts are important for conservation, management, and
evaluating how well current models characterize natural systems. While the
number of forecasts for biodiversity is increasing, there is little information
available on how well these forecasts work. Most biodiversity forecasts are not
evaluated to determine how well they predict future diversity, fail to account
for uncertainty, and do not use time-series data that captures the actual
dynamics being studied. We addressed these limitations by using best practices
to explore our ability to forecast the species richness of breeding birds in
North America. We used hindcasting to evaluate six different modeling approaches
for predicting richness. Hindcasts for each method were evaluated annually for a
decade at `r numbers$N_predict_sites` sites distributed throughout the
continental United States. While each model could explain most of the variance
in richness, none of them consistently outperformed a baseline model that
predicted constant richness at each site. In particular, we found no evidence
that current methods (such as species distribution models) can successfully turn
spatial data into useful temporal predictions about biodiversity at decadal
time-scales. The best practices implemented in this study directly influence the
forecasts, the relative performance of different modeling approaches, and the
conclusions about the current state of biodiversity forecasting . To facilitate
the rapid improvement of biodiversity forecasts, we emphasize the value of
specific best practices in making forecasts and evaluating forecasting methods.

@@

Forecasting is an emerging imperative in ecology; as such, the field needs to
develop and follow best practices for conducting and evaluating ecological
forecasts [@clark2001]. We have used a number of these practices (Box 1) in a
single study that builds and evaluates forecasts of biodiversity in the form of
species richness. The results of this effort are both promising and humbling.
When comparing forecasts across sites, many different approaches to forecasting
produce reasonable forecasts (Figure 3). If a site is predicted to have a high
number of species in the future, relative to other sites, it generally does.
However, none of the methods evaluated reliably determined how site-level
richness changes over time (Figure 6), which is generally the stated purpose of
these forecasts. As a result, baseline models, which did not attempt to
anticipate changes in richness over time, generally provided the best forecasts
for future biodiversity. While this study was based on a single data set, its
results are consistent with a growing literature on the limits of ecological 
forecasting, as discussed below.

The most commonly used methods for forecasting future biodiversity, SDMs and
macroecological models, both produced worse forecasts than time-series models
and simple baselines. This weakness suggests that predictions about future
biodiversity change should be viewed with skepticism unless the underlying
models have been validated temporally, via hindcasting. Since site-level
richness is relatively stable, spatial validation is not enough: a model can
have high accuracy across spatial gradients without being able to predict
changes over time. This gap between spatial and temporal accuracy is known to be
important for species-level predictions [@rapacciuolo2012; @oedekoven2017]; our
results indicate that it is substantial for higher-level patterns like richness
as well. SDMs' poor temporal predictions are particularly sobering, as these
models have been one of the main foundations for estimates of the predicted loss
of biodiversity to climate change over the past decade or so [@thomas2004;
@thuiller2011; @urban2015]. Our results also highlight the importance of
comparing multiple modeling approaches when conducting ecological forecasts, and
in particular, the value of comparing results to simple baselines to avoid
over-interpreting the information present in these forecasts [Box 1].
Disciplines that have more mature forecasting cultures often do this by
reporting "forecast skill", i.e., the improvement in the forecast relative to a
simple baseline [@jolliffe2003]. We recommend following the example of @ye2015
and adopting this approach in future ecological forecasting research.

When comparing different methods for forecasting our results demonstrate the
importance of considering uncertainty [Box 1; @clark2001; @dietz2016]. Previous
comparisons between stacked SDMs and macroecological models reported that the
methods yielded equivalent results for forecasting diversity [@algar2009;
@distler2015]. While our results support this equivalence for point estimates,
they also show that stacked SDMs dramatically underestimate the range of
possible outcomes; after ten years, more than a third of the observed richness
values fell outside the stacked SDMs' 95% prediction intervals. Consistent with
@harris2015 and @warton2015, we found that JSDMs' wider prediction intervals
enabled them to avoid this problem. Macroecological models appear to share this
advantage, while being considerably easier to implement.
